Galois Characterization of Function Fields of \\ Transcendence Degree $1$ over Number Fields

### 2017-05-20 Master Thesis/synopsis.tex
Add more chapter comments

This section we recall facts from valuation theory.

@@

Using the bijection $\operatorname{Spv}_\text{d}^{(2)}(K)\to \operatorname{Spv}_\text{d}^{(2)}(K')$ above we want to establish a bijection $\operatorname{Spv}_\text{rd}^{(2)}(K)\to \operatorname{Spv}_\text{rd}^{(2)}(K')$ in order to establish a bijection between the defectless rank-$1$ valuations:

@@

In this section we shall show that $\sigma: G_K\to G_{K'}$ induces group isomorphisms $K'^\times/n\to K^\times/n$ for all $n\in\mathbb{N}$.

@@

\begin{remark}
Consider $\overline{K}'' = \overline{K}'\otimes_{\overline{k'}}\overline{k}$ via the isomorphism $\varphi: \overline{k'}\to \overline{k}$. If we replace $\sigma: G_K\to G_{K'}$ with the composition
\[ \begin{tikzcd}
G_K \ar[r, "\sigma"] & G_{K'} \ar[r] & G_{K'}
\end{tikzcd} \]
where the second map is induced by the isomorphism $\overline{K''}\to\overline{K'},\ x\otimes y\mapsto \varphi^{-1}(y)\cdot x$, we may assume that $k=k'$, $\overline{k} = \overline{k'}$ and that $\varphi:\overline{k}\to\overline{k}$, $\sigma: G_k\to G_k$ are identities.
\end{remark}

\paragraph{} Let $n\in\mathbb{N}$. In the previous remark, we have replaced $\sigma$, so that the following diagram commutes now:
\[ \begin{tikzcd}
G_K \ar[r]\ar[d, "\sigma"'] & G_k\\
G_{K'}\ar[ru]
\end{tikzcd} \]
The operation of $G_K$ and $G_{K'}$ on $\mu_n$ factors through $G_k$, thus we have $\sigma(g)\xi = g\xi$ for all $\xi\in\mu_n$ and $g\in G_K$. Therefore, for all intermediate fields $L$ of $\overline{K}/K$, $\sigma$ induces an isomorphism:
\[ \varphi_n:L'^\times/n \cong\mathrm{H}^1(L', \mu_n) \to \mathrm{H}^1(L,\mu_n) \cong L^\times/n \]

@@

\begin{proof}
\begin{enumerate}[(i)]
\item Consider the absolute inertia field $L = K_P^\text{t}$ of $K_P$ and let $w$ be the unique extension of $v_P$ in $L$. By Theorem \ref{theorem-absoluteinertia} (i), $w$ is discrete as well with $k(w) = k(v_P)^\text{s}$. We have the following exact sequence:
\[ 1 \longrightarrow U_w/n \longrightarrow L^\times /n\longrightarrow \mathbb{Z}/n\mathbb{Z}\longrightarrow 0 \]
Since $k(w)$ is separably closed, with $\operatorname{char} k(w)=0$ and Hensel's lemma we can see that $U_w$ is a divisible group, hence we get an isomorphism $L^\times /n\cong \mathbb{Z}/n\mathbb{Z}$ induced by $w$. Therefore, the canonical map $K^\times/n\to L^\times/n$ can be identified with the map $K^\times/n \to\mathbb{Z}/n\mathbb{Z}$ induced by $v_P$. Now the following commutative diagram with exact columns gives us the result:
\[ \begin{tikzcd}
1 \ar[d] & 1 \ar[d]\\
U_{P'}/n \ar[r, dashed]\ar[d] & U_P/n\ar[d]\\
K'^\times /n \ar[r, "\varphi_n"]\ar[d] & K^\times/n\ar[d]\\
L'^\times /n \ar[r, "\varphi_n"'] &  L^\times /n
\end{tikzcd}\]
Note that $(K^\text{t}_P)' = K'^{\text{t}}_{P'}$ as proven in Lemma \ref{3.2}.
\item Consider the field extension $K_P/K$. By Remark \ref{remark-localphi-extension}, without loss of generality, we can assume $K=K_P$. Since $\operatorname{char}k(P)=0$, by Hensel's lemma, $r_P/n$ is an isomorphism and the composition
\[\begin{tikzcd}
k(P)^\times / n\ar[rr, "(r_P/n)^{-1}"] && U_P/n \ar[rr, hook] && K_P^\times/n
\end{tikzcd}\]
coincides with the inflation $\operatorname{inf}: \mathrm{H}^1(k(P), \mu_n) \to \mathrm{H}^1(K_P,\mu_n)$. The same holds for $P'$ due to Lemma \ref{3.2}. Therefore, the assertion follows from the commutativity of the following diagram:
\[ \begin{tikzcd}
k(P')^\times/n \ar[r, "\cong"] \ar[d, "\varphi_P/n"'] & \mathrm{H}^1(k(P'),\mu_n) \ar[r, "\mathrm{inf}"] \ar[d]& \mathrm{H}^1(K'_{P'}, \mu_n) \ar[d, "\varphi_n"] \\
k(P)^\times/n \ar[r, "\cong"'] & \mathrm{H}^1(k(P),\mu_n)  \ar[r, "\mathrm{inf}"'] & \mathrm{H}^1(K_{P}, \mu_n)
\end{tikzcd}  \qedhere\]
\end{enumerate}
\end{proof}

@@

@@

\begin{definition}\label{5.1}
An abelian group $A$ is called \textit{free-by-finite} if its torsion subgroup $A_\text{tor}$ is finite and the quotient $A/A_\text{tor}$ is free abelian. $A$ is free-by-finite if and only if $A$ is the direct sum of a finite abelian group and a free abelian group.
\end{definition}

@@

\begin{proof}
\begin{enumerate}[(i)]
\item Let $B$ be free-by-finite and $K = \ker(f)$. Considering $A/K$ as a subgroup of $B$, we see that $A/K$ is free-by-finite. Since $(A/K)_\text{tor}$ and $K$ are finite, $K' = \{a\in A\mid na\in K \text{ for some }n\in\mathbb{N}\}$ is finite as well, in particular $A_\text{tor}$ is finite. In fact $K'=A_\text{tor}$, since $K\subset A_\text{tor}$. Now consider the surjection $A\to A/K\to (A/K)/(A/K)_\text{tor}$. This map has kernel $K'$, thus inducing an isomorphism from $A/A_\text{tor}$ to a free abelian group.

Now let $A$ be free-by-finite, i.e. $A = A_\text{tor} \oplus F$ for some free abelian group $F$, and $K=\ker(f)$. Since $K\subset A$ is finite, $K$ is actually contained in $A_\text{tor}$. Therefore, we see that $A/K$ is free by finite as well, and by extension, the image $I$ of $f$. Set $m = \#B/I$ and consider the canonical map $B\stackrel{\cdot m}{\to} I \to I/I_\text{tor}$. This map has kernel $\{b\in B \mid mb \in I_\text{tor}\} = B_\text{tor}$, making $B/B_\text{tor}$ isomorphic to a subgroup of the free abelian group $I/I_\text{tor}$. Now both $I_\text{tor}$ and $B_\text{tor}/I_\text{tor} \subset B/I$ are finite, hence $B_\text{tor}$ is finite as well.
\item The exact sequence induces the exact sequence $0 \longrightarrow A'_\text{tor} \longrightarrow A_\text{tor} \longrightarrow A''_\text{tor}$. Since the surrounding groups are finite, $A_\text{tor}$ is finite as well. Using (i) we see that it suffices to show that $A/A_\text{tor}$ is free-by-finite.

@@

\begin{lemma}\label{5.3}
Let $F$ be a number field. Then:
\begin{enumerate}[(i)]
\item $F^\times$ is free-by-finite.
\item Let $F_1,\ldots,F_n$ be finite extensions of $F$. Then $\big(\prod_{i=1}^{n}F_i^\times\big)/F^\times$ is free-by-finite, where $F^\times$ is considered as a subgroup of $\prod_{i=1}^{n}F_i^\times$ via the diagonal embedding.
\end{enumerate}
\end{lemma}

@@

\section{Proof of Theorem}

>>>>

In this section we shall recall facts from valuation theory.

@@

We will see that the bijection $\operatorname{Spv}_\text{d}^{(2)}(K)\to \operatorname{Spv}_\text{d}^{(2)}(K')$ above restricts to a bijection $\operatorname{Spv}_\text{rd}^{(2)}(K)\to \operatorname{Spv}_\text{rd}^{(2)}(K')$. Using this, we shall establish a bijection between the defectless valuations of rational rank $1$:

@@

In this section we shall show that $\sigma: G_K\to G_{K'}$ induces group isomorphisms $K'^\times/n\to K^\times/n$ and $U_{v'}/n\to U_v/n$ for all $n\in\mathbb{N}$ and defectless valuations $v$ of rational rank $1$ with $\operatorname{char}k(v) =0$.

@@

\begin{remark}
Consider $\overline{K}'' = \overline{K}'\otimes_{\overline{k'}}\overline{k}$ via the isomorphism $\varphi: \overline{k'}\to \overline{k}$. If we replace $\sigma: G_K\to G_{K'}$ with the composition
\[ \begin{tikzcd}
G_K \ar[r, "\sigma"] & G_{K'} \ar[r, "\phi"] & G_{K'}
\end{tikzcd} \]
where $\phi$ is induced by the isomorphism $\overline{K''}\to\overline{K'},\ x\otimes y\mapsto \varphi^{-1}(y)\cdot x$, we may assume that $k=k'$, $\overline{k} = \overline{k'}$ and that $\varphi:\overline{k}\to\overline{k}$, $\sigma: G_k\to G_k$ are identities. Now the following diagram commutes:
\[ \begin{tikzcd}
G_K \ar[r]\ar[d, "\sigma"'] & G_k\\
G_{K'}\ar[ru]
\end{tikzcd} \]
\end{remark}

\paragraph{} Let $n\in\mathbb{N}$. The operation of $G_K$ and $G_{K'}$ on $\mu_n$ factors through $G_k$, thus we have $\sigma(g)\xi = g\xi$ for all $\xi\in\mu_n$ and $g\in G_K$. Therefore, for all intermediate fields $L$ of $\overline{K}/K$, $\sigma$ induces an isomorphism:
\[ \varphi_n:L'^\times/n \cong\mathrm{H}^1(L', \mu_n) \to \mathrm{H}^1(L,\mu_n) \cong L^\times/n \]

@@

\begin{proof}
\begin{enumerate}[(i)]
\item Consider the absolute inertia field $L = K_P^\text{t}$ of $K_P$ and let $w$ be the unique extension of $v_P$ in $L$. By Theorem \ref{theorem-absoluteinertia} (i), $w$ is discrete as well and we have $k(w) = k(v_P)^\text{s}$. Observe the following exact sequence:
\[ 1 \longrightarrow U_w/n \longrightarrow L^\times /n\longrightarrow \mathbb{Z}/n\mathbb{Z}\longrightarrow 0 \]
Since $k(w)$ is separably closed, with $\operatorname{char} k(w)=0$ and Hensel's lemma we can see that $U_w$ is a divisible group, hence we get an isomorphism $L^\times /n\cong \mathbb{Z}/n\mathbb{Z}$ induced by $w$. Therefore, the canonical map $K^\times/n\to L^\times/n$ can be identified with the map $K^\times/n \to\mathbb{Z}/n\mathbb{Z}$ induced by $v_P$. Now the following commutative diagram with exact columns gives us the result:
\[ \begin{tikzcd}
1 \ar[d] & 1 \ar[d]\\
U_{P'}/n \ar[r, dashed]\ar[d] & U_P/n\ar[d]\\
K'^\times /n \ar[r, "\varphi_n"]\ar[d] & K^\times/n\ar[d]\\
L'^\times /n \ar[r, "\varphi_n"'] &  L^\times /n
\end{tikzcd}\]
Note that $(K^\text{t}_P)' = K'^{\text{t}}_{P'}$ as proven in Lemma \ref{3.2}.
\item Consider the field extension $K_P/K$. By Remark \ref{remark-localphi-extension}, without loss of generality, we can assume $K=K_P$. Since $\operatorname{char}k(P)=0$, by Hensel's lemma, $r_P/n$ is an isomorphism and the composition
\[\begin{tikzcd}
k(P)^\times / n\ar[rr, "(r_P/n)^{-1}"] && U_P/n \ar[rr, hook] && K_P^\times/n
\end{tikzcd}\]
coincides with the inflation $\operatorname{inf}: \mathrm{H}^1(k(P), \mu_n) \to \mathrm{H}^1(K_P,\mu_n)$. The same holds for $P'$ due to Lemma \ref{3.2}. Therefore, the assertion follows from the commutativity of the following diagram:
\[ \begin{tikzcd}
k(P')^\times/n \ar[r, "\cong"] \ar[d, "\varphi_P/n"'] & \mathrm{H}^1(k(P'),\mu_n) \ar[r, "\mathrm{inf}"] \ar[d]& \mathrm{H}^1(K'_{P'}, \mu_n) \ar[d, "\varphi_n"] \\
k(P)^\times/n \ar[r, "\cong"'] & \mathrm{H}^1(k(P),\mu_n)  \ar[r, "\mathrm{inf}"'] & \mathrm{H}^1(K_{P}, \mu_n)
\end{tikzcd}  \qedhere\]
\end{enumerate}
\end{proof}

@@

Before we start proving Pop's Theorem \ref{thm:main-result}, we need some tools that will help us constructing a group isomorphism $K'^\times\to K^\times$ from the modulo $n$ isomorphisms $K'^\times/n\to K^\times/n$.

@@

\begin{definition}\label{5.1}
Let $A$ be an abelian group. We have the following short exact sequence:
\[0\longrightarrow A_\text{tor} \longrightarrow A\longrightarrow A/A_\text{tor}\longrightarrow 0 \]
$A$ is called \textit{free-by-finite} if its torsion subgroup $A_\text{tor}$ is finite and the quotient $A/A_\text{tor}$ is free abelian. If $A$ is free-by-finite, the above sequence splits and $A$ is the direct sum of a finite abelian group and a free abelian group. It's evident that the converse also holds.
\end{definition}

@@

\begin{proof}
\begin{enumerate}[(i)]
\item Let $B$ be free-by-finite and $K = \ker(f)$. Considering $A/K$ as a subgroup of $B$, we see that $A/K$ is free-by-finite. Since $(A/K)_\text{tor}$ and $K$ are finite, $K' = \{a\in A\mid na\in K \text{ for some }n\in\mathbb{N}\}$ is finite as well, in particular $A_\text{tor}\subset K'$ is finite. In fact $K'=A_\text{tor}$, since $K\subset A_\text{tor}$. Now consider the surjection $A\to A/K\to (A/K)/(A/K)_\text{tor}$. This map has kernel $K'$, thus inducing an isomorphism from $A/A_\text{tor}$ to a free abelian group.

Now let $A$ be free-by-finite, i.e. $A = A_\text{tor} \oplus F$ for some free abelian group $F$, and $K=\ker(f)\subset A_\text{tor}$. We see that $A/K$ is free by finite as well, and by extension, the image $I$ of $f$. Set $m = \#B/I$ and consider the canonical map $B\stackrel{\cdot m}{\to} I \to I/I_\text{tor}$. This map has kernel $\{b\in B \mid mb \in I_\text{tor}\} = B_\text{tor}$, making $B/B_\text{tor}$ isomorphic to a subgroup of the free abelian group $I/I_\text{tor}$. Now both $I_\text{tor}$ and $B_\text{tor}/I_\text{tor} \subset B/I$ are finite, hence $B_\text{tor}$ is finite as well.
\item The exact sequence induces the exact sequence:
\[0 \longrightarrow A'_\text{tor} \longrightarrow A_\text{tor} \longrightarrow A''_\text{tor}\] 
Since the surrounding groups are finite, $A_\text{tor}$ is finite as well. Using (i) we see that it suffices to show that $A/A_\text{tor}$ is free-by-finite.

@@

\begin{lemma}\label{5.3}
Let $F$ be a number field. Then:
\begin{enumerate}[(i)]
\item $F^\times$ is free-by-finite.
\item Let $F_1,\ldots,F_n$ be finite extensions of $F$. Then 
\[\Big(\prod_{i=1}^{n}F_i^\times\Big)/F^\times\] 
is free-by-finite, where $F^\times$ is considered as a subgroup of $\prod_{i=1}^{n}F_i^\times$ via the diagonal embedding.
\end{enumerate}
\end{lemma}

@@

\section{Proof of Pop's Theorem}
